Stability in human interaction networks: primitive typology of vertex, prominence of measures and activity statistics

### 2015-04-02 paper.tex
feito basico do seccionamento de erdos

@@

 Next subsection exposes related work. Section~\ref{sec:data} details data while Section~\ref{sec:carac} is dedicated to the characterization methods. Section~\ref{sec:results} holds results and discussions, leading to Section~\ref{sec:conc} for conclusions and further work. Appendix~\ref{ap:ded} develops simple graph results used in the methods, Appendix~\ref{scripts} exposes access to data and scripts, and Appendix~\ref{ap:circular} presents the circular statistics (or directional statistics) used in characterizing time activity. Detailed tables and figures are bundled in Support Information, and are cited upon need.

@@

The evolution of human interaction networks was addressed with a community focus, where the direction of edges was not taken into account~\cite{barabasiEvo}. In another article, two topologically different networks emerged from human interaction networks, depending on the frequency of interactions, which could either be a generalized power law or an exponential connectivity distribution~\cite{barabasiTopologicalEv}. In email list networks, scale-free properties were verified~\cite{bird}, and different linguistic traces were related to weak and strong ties~\cite{GMANE2}.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/fser_}
    \caption{Degree distribution of scale-free and Erd\"os-R\'enyi ideal networks. The latter has more
        intermediary vertices, while the former has more peripheral and hub vertices. Sector borders are
        given by the two intersections $k_1$ and $k_2$ of the connectivity distributions. Characteristic degrees
    are in compact intervals of degree: $[0,k_1]$, $(k_1,k_2]$, $(k_2,k_{max}]$ for the Erd\"os sectors (periphery, intermediary and hubs).}
    \label{fig:setores}
\end{figure}

@@

If strength $s$ is used for comparison, $P$ remains the same, but $P(\kappa_i)$ with $\kappa_i=\frac{s_i}{\overline{w}}$ should be used for comparison, with $\overline{w}=2\frac{z}{\sum_is_i}$ the average weight of an edge and $s_i$ the strength of vertex $i$. For in and out degrees and strengths, comparison should be made with $\kappa_i=2k_i^{in}$, $\kappa_i=2k_i^{out}$, $\kappa_i=2\frac{s_i^{in}}{\overline{w}}$ and $\kappa_i=2\frac{s_i^{out}}{\overline{w}}$. Results of these criteria for network segmentation are discussed in Section~\ref{subsec:pih} and exhibited in Figures~\ref{fig:cpp10000} to~\ref{fig:lad50_} of Appendix~\ref{figures}.

@@

These compound criteria, and the simplification of possibilities listed above (exclusivist/inclusivist, criterion/cascade/externals), can be formalized in strict mathematical terms, but this was considered out of the scope of the present article. Important here is to notice that compound criteria can be used to examine network sections in the case of a low number of messages, such as in the last figures of Appendix~\ref{figures}.

@@

\subsubsection{Topology metrics}\label{measures}

The topology of the networks was characterized with a small selection of the most basic and fundamental measurements for each vertex, as follows:

\begin{itemize}
    \item Degree $d_i$: number of edges linked to vertex $i$.
    \item In-degree $d_i^{in}$: number of edges ending at vertex $i$.
    \item Out-degree $d_i^{out}$: number of edges departing from vertex $i$.
    \item Strength $s$: sum of weights of all edges linked to vertex $i$.
    \item In-strength $s_i^{in}$: sum of weights of all edges ending at vertex $i$.
    \item Out-strength $s_i^{out}$: sum of weights of all edges departing from vertex $i$.
    \item Clustering coefficient $cc_i$: fraction of pairs of neighbors of $i$ that are linked.  The standard clustering coefficient for undirected graphs was used.
    \item Betweenness centrality $bt_i$: fraction of geodesics that contain the vertex $i$. Betweenness centrality index considered directions and weight, as specified in~\cite{faster}.
\end{itemize}

@@

\begin{itemize}
    \item Asymmetry: $asy_i=\frac{d_i^{in}-d_i^{out}}{d_i}$.
    \item Mean of asymmetry of edges: $\mu_i^{asy}=\frac{\sum_{j\in J_i} e_{ji}-e_{ij}}{|J_i|=d_i}$. Where $e_{xy}$ is 1 if there is and edge from $x$ to $y$, $0$ otherwise. $J_i$ is the set of neighbors of vertex $i$, and $|J_i|=d_i$ is the number of neighbors of vertex $i$.
    \item Standard deviation of asymmetry of edges: $\sigma_i^{asy}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{asy} -(e_{ji}-e_{ij}) ]^2  }{d_i}  }$.
    \item Disequilibrium: $dis_i=\frac{s_i^{in}-s_i^{out}}{s_i}$.
    \item Mean of disequilibrium of edges: $\mu_i^{dis}=\frac{\sum_{j \in J_i}\frac{w_{ji}-w_{ij}}{s_i}}{d_i}$, where $w_{xy}$ is the weight of edge $x\rightarrow y$ and zero if there is no such edge.
    \item Standard deviation of disequilibrium of edges: $\sigma_i^{dis}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{dis}-\frac{(w_{ji}-w_{ij})}{s_i}]^2}{d_i}}$.
\end{itemize}

   \subsection{Evolution of the networks}
   The evolution of the networks was observed within a fixed number of messages, the window size $ws$. This same number of contiguous messages is considered with different shifts in the message timeline to obtain snapshots. Each snapshot was used to both do the Erd\"os sectioning (see Section~\ref{sectioning}) and take topological measures (see Section~\ref{measures}). Furthermore, this approach to evolution was used for visualization (see Section~\ref{sec:viz}) and for typological thinking (see Section~\ref{subsec:typ}).
The $ws$ used were 50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000. Within a same $ws$, the number of vertices and edges vary in time, as do other network characteristics.

@@

    \subsection{Typological deepening}\label{subsec:typ}
There are other ways to split and characterize networks. To point a common example, the center of the network is defined as all the nodes whose maximum distance to any other node is the radius (the radius is the minimum maximum distance to all vertices , i.e. the radius is the minimum eccentricity). 
In the same framework, the periphery (as opposed to the center) consists of the nodes whose maximum distance to any node is the diameter (diameter being the maximum geodesic on the network). Accordingly, the intermediary sector can be defined as the nodes that are not in the center or in the periphery. Interestingly, in the email networks analyzed, with such criteria, the center can often be a factor of 4 times larger than the periphery and the intermediary group often exceeds 90\% of the nodes~\cite{networkx}.

Models of human dynamics can be used to predict and classify activity. In this case, agent activity is commonly considered a Poisson process, as a consequence of the randomly distributed events in time. Even so, evidence-based models suggests that human activity patterns follow non-Poisson statistics, characterized by a long tail of inactivity with bursts of rapidly occurring events~\cite{barabasiHumanDyn,barabasiPhone}. Emails are reported as having a heavy tailed distribution with $\alpha=1$, together with web browsing and library loans~\cite{barabasiHumanDyn}.

%Influenced by Social and Psychoanalytic Theories, Adorno et all applied a %questionnaire to individuals, from which they reached a position in the the %``F Scale'', to verify etnocentric, conservatory and antidemocratic %trends~\cite{adorno}. From psychoanalitic interviews and the F Scale, they %derived a typology, which gathers prejudice-inclined traces in personality. %Both, low and high scores are considered with prejudicial traces. This %typology has nine authoritarian types, the six types with high score in the F %Scale: surface resentment, conventional, authoritarian, rebel and psychopath, %crank, manipulative; and three of the five types with low score in the F %Scale: rigid, protesting, impulsive, easygoing, genuine liberal. Each side of %the dipole has a rank of intensity that increases as the order written above.

%    \subsection{Other analisys of interest}
%
%        \subsubsection{Textual content}
%        \subsubsection{Geographical localization}

@@

Remarkable features from the analysis of the four email lists are:
\begin{itemize}
    \item The activity along time is practically the same for all lists (Section~\ref{constDisc}).
    \item The fraction of participants in each connective sector is stable and can be observed even with very few messages (Section~\ref{subsec:pih}).
    \item The measures combine in principal components in the same way and with symmetry related measures presenting more dispersion than clustering coefficient (Section~\ref{prevalence}).
    \item Typology speculations are immediate from results (Section~\ref{sec:pty}).
\end{itemize}

Furthermore, the Appendix holds tables and figures from which such results can be observed.

\subsection{Constancy and discrepancy of activity along time}\label{constDisc}
One remarkable feature from the analysis is that the activity along time is practically the same for all lists.

\subsubsection{Seconds and minutes}
The incidence of messages at each second of a minute and at each minute of an hour is compatible with uniform distribution simulations\footnote{Numpy version 1.6.1, ``random.randint'' function, was used for simulations.}. Messages were slightly more evenly distributed in all lists: for both seconds and minutes  $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$. Simulations reach these values  but have in average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$.

@@

\subsection{Scalable fat-tail structure: constancy of membership fractions in the connective sectors}\label{subsec:pih}
There is a concentration of hub activity and of vertex with few connections. Table~\ref{autores} is dedicated to exposing this well known and expected distribution of activity among participants.

@@

Regarding topological aspects of interaction networks, further work should inspect other measures in each of the three connective sectors: hubs, intermediary and periphery.
Observance of attributes with greater contribution to principal components of LDA should reveal best chances to present these three sections as clusters in the network measurements space. Another possibility, specially for a brute-force characterization of such sectors, is to remove vertices with degree close to $k_1$ or $k_2$ depicted in figure~\ref{fig:setores}. The subtraction $\widetilde{P}(k)-P(k)$ (see Section~\ref{subsec:typ}) results in two positive clusters for periphery and hubs, and a negative cluster for intermediary vertices. This might support classification of the three sectors by clustering, a more traditional approach to classification.

@@

\subsection{Third party libraries and software}
The programming framework used 
is mainly Python-based, with emphasis on usual
scientific tools. More specifically,
scripts where written for 2.7.3 version of Python,
with the following third party libraries: Numpy, Pylab/Matplotlib, NetworkX, IGraph.
Behind the scenes, Graphviz is accessed via PyGraphviz to make network drawings.
\section{Circular statistics}\label{ap:circular}
\section{Tables}\label{sectables}
\clearpage
\subsection{PCA tables}\label{sec:pcat}

>>>>

Typologies are the canon of scientific literature for classification of human agents~\cite{typCanon}. Consequently, it was considered central to benefit from human typologies. This is a complex matter carefully addressed in methodology, results and appendix.

@@

Next subsection exposes related work. Section~\ref{sec:data} details data while Section~\ref{sec:carac} is dedicated to the characterization methods. Section~\ref{sec:results} holds results and discussions, leading to Section~\ref{sec:conc} for conclusions and further work. Appendix~\ref{ap:ded} develops simple graph results used in the methods, Appendix~\ref{scripts} exposes access to data and scripts, Appendix~\ref{ap:circular} presents the circular statistics (or directional statistics) used in characterizing time activity, and Appendix~\ref{ap:typ} further introduces the typological approach. Detailed tables and figures are bundled in Support Information, and are cited upon need.

@@

The evolution of human interaction networks was addressed with a community focus, where the direction of edges was not taken into account~\cite{barabasiEvo}. In another article, two topologically different networks emerged from human interaction networks, depending on the frequency of interactions, which can either be a generalized power law or an exponential connectivity distribution~\cite{barabasiTopologicalEv}. In email list networks, scale-free properties were reported with $\alpha=1$~\cite{bird} (as are web browsing and library loans~\cite{barabasiHumanDyn}), and different linguistic traces were related to weak and strong ties~\cite{GMANE2}.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/fser_}
    \caption{Degree distribution of scale-free and Erd\"os-R\'enyi ideal networks. The latter has more
        intermediary vertices, while the former has more peripheral and hub vertices. Sector borders are
        given by the two intersections $k_\Leftarrow$ and $k_\Rightarrow$ of the connectivity distributions. Characteristic degrees
    are in compact intervals of degree: $[0,k_\Leftarrow]$, $(k_\Leftarrow,k_\Rightarrow]$, $(k_\Rightarrow,k_{max}]$ for the Erd\"os sectors (periphery, intermediary and hubs).}
    \label{fig:setores}
\end{figure}

@@

If strength $s$ is used for comparison, $P$ remains the same, but $P(\kappa_i)$ with $\kappa_i=\frac{s_i}{\overline{w}}$ should be used for comparison, with $\overline{w}=2\frac{z}{\sum_is_i}$ the average weight of an edge and $s_i$ the strength of vertex $i$. For in and out degrees and strengths, comparison of the real network should be made with:
\begin{equation}
\hat{P}(k^{way})=\binom{N-1}{k^{way}}p_e^k(1-p_e)^{N-1-k^{way}}
\end{equation}

\noindent where \emph{way} can be \emph{in} or \emph{out}. In and out strengths are divided by $\overline{w}$ and compared also using $\hat{P}$. Notice that $p_e$ remains the same, as each edge yields an incoming (or outgoing) edge, and there are at most $N(N-1)$ incoming (or outgoing) edges, thus $p_e=\frac{z}{N(N-1)}$.

In other words, let $\sigma$ and $\phi$ be integers in the intervals $1 \leq \sigma \leq 6$, $1 \leq \phi \leq 3$, and the basic six Erd\"os sectioning possibilities $\{E_{\sigma}\}$ have Erd\"os sectors $E_{\sigma}= \{e_{\sigma, \phi} \}$ defined as:

\begin{alignat}{3}
e_{\sigma,1}&=\{\;i\;|\;\overline{k}_{\sigma,\Leftarrow}\geq&&\overline{k}_{\sigma,i}\} \\
e_{\sigma,2}&=\{\;i\;|\;\overline{k}_{\sigma,\Leftarrow}<\;&&\overline{k}_{\sigma,i}\leq\overline{k}_{\sigma,\Rightarrow}\} \\ 
e_{\sigma,3}&=\{\;i\;|\;&&\overline{k}_{\sigma,i}<\overline{k}_{\sigma,\Rightarrow}\}
\end{alignat}

\noindent where $\{\overline{k}_{\sigma,i}\}$ is defined as:

\begin{equation}
\begin{split}
\overline{k}_{1,i}&=k_i \\
\overline{k}_{2,i}&=k_i^{in} \\
\overline{k}_{3,i}&=k_i^{out} \\
\overline{k}_{4,i}&=\frac{s_i}{\overline{w}} \\
\overline{k}_{5,i}&=\frac{s_i^{in}}{\overline{w}} \\
\overline{k}_{6,i}&=\frac{s_i^{out}}{\overline{w}} \\
\end{split}
\end{equation}

Results of these criteria for network segmentation are discussed in Section~\ref{subsec:pih} and exhibited in Figures~\ref{fig:cpp10000} to~\ref{fig:lad50_} of Appendix~\ref{figures}.

@@

These compound criteria, and the simplification of all the compound possibilities as listed above (into exclusivist/inclusivist criterion/cascade/externals), can be formalized in strict mathematical terms, but this was considered out of the scope of the present article. Notice that the exclusivist cascade is the same sectioning that an inclusivist cascade from periphery to hubs (inverted order of precedence). Important here is to notice that compound criteria can be used to examine network sections in the case of a low number of messages, such as in the last figures of Appendix~\ref{figures}.

@@

\subsubsection{Topology metrics for Principal Component Analysis}\label{measures}

The topology of the networks was observed by Principal Component Analysis (PCA~\cite{pca}) outcomes
with a small selection of the most basic and fundamental measurements for each vertex:

\begin{itemize}
    \item Degree     $k_i$: number of edges linked to vertex $i$.
    \item In-degree  $k_i^{in}$: number of edges ending at vertex $i$.
    \item Out-degree $k_i^{out}$: number of edges departing from vertex $i$.
    \item Strength $s$: sum of weights of all edges linked to vertex $i$.
    \item In-strength $s_i^{in}$: sum of weights of all edges ending at vertex $i$.
    \item Out-strength $s_i^{out}$: sum of weights of all edges departing from vertex $i$.
    \item Clustering coefficient $cc_i$: fraction of pairs of neighbors of $i$ that are linked.  The standard clustering coefficient for undirected graphs was used.
    \item Betweenness centrality $bt_i$: fraction of geodesics that contain the vertex $i$. Betweenness centrality index considered directions and weight, as specified in~\cite{faster}.
\end{itemize}

@@

\begin{itemize}
    \item Asymmetry: $asy_i=\frac{k_i^{in}-k_i^{out}}{k_i}$.
    \item Mean of asymmetry of edges: $\mu_i^{asy}=\frac{\sum_{j\in J_i} e_{ji}-e_{ij}}{|J_i|=k_i}$. Where $e_{xy}$ is 1 if there is and edge from $x$ to $y$, $0$ otherwise. $J_i$ is the set of neighbors of vertex $i$, and $|J_i|=k_i$ is the number of neighbors of vertex $i$.
    \item Standard deviation of asymmetry of edges: $\sigma_i^{asy}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{asy} -(e_{ji}-e_{ij}) ]^2  }{k_i}  }$.
    \item Disequilibrium: $dis_i=\frac{s_i^{in}-s_i^{out}}{s_i}$.
    \item Mean of disequilibrium of edges: $\mu_i^{dis}=\frac{\sum_{j \in J_i}\frac{w_{ji}-w_{ij}}{s_i}}{k_i}$, where $w_{xy}$ is the weight of edge $x\rightarrow y$ and zero if there is no such edge.
    \item Standard deviation of disequilibrium of edges: $\sigma_i^{dis}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{dis}-\frac{(w_{ji}-w_{ij})}{s_i}]^2}{k_i}}$.
\end{itemize}

\subsection{Evolution of the networks}
   The evolution of the networks was observed within a fixed number of messages, the window size $ws$. This same number of contiguous messages is considered with different shifts in the message timeline to obtain snapshots. Each snapshot was used to both do the Erd\"os sectioning (see Section~\ref{sectioning}) and to perform PCA with topological measures (see Section~\ref{measures}). Furthermore, this approach to evolution was used for visualization (see Section~\ref{sec:viz}) and for typological thinking (see Section~\ref{subsec:typ}).
The $ws$ used were 50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000. Within a same $ws$, the number of vertices and edges vary in time, as do other network characteristics.

@@

\subsection{Typology speculation}\label{subsec:typ}
Qualitative, typological speculations were the result of all the methodology. More specifically, the Erd\"os sectors was regarded as yielding a \emph{primitive typology} of human agents in social contexts with at least dozens of participants. 
Visualization and data mining were efficient in fueling speculations about the qualities of each type. 
The stability found in principal components of topological measures suggested that this primitive typology is general for (practically) all human contexts.
Also, the first author is a developer with cultural interests and has been directly and indirectly part of these communities for more than a decade, which gave further safety about assumptions. Care was taken not to regard this acquaintanceship as primary source or as definitive evidence.
The interested reader should see Appendix~\ref{ap:typ} for further context, 
as it might be considered audacious to bridge from a physics-based, 
quantitative classification to a qualitative and speculative approach.

@@

Remarkable features from the analysis of the four email lists are:
\begin{itemize}
    \item The activity along time is practically the same for all lists (Section~\ref{constDisc}) and exhibited noticeable patterns.
    \item The fraction of participants in each Erd\"os sector is stable and can be observed even with very few messages (Section~\ref{subsec:pih}).
    \item The measures combine into principal components in the same way for all lists and all snapshots. 
	    Furthermore, symmetry related measures, defined in this article, present more dispersion than the usual clustering coefficient (Section~\ref{prevalence}).
    \item Typology speculations are immediate from results (Section~\ref{sec:pty}).
\end{itemize}

The reminder of this section contains a detailed discussion of these results and
Support Information holds tables and figures from which such results can be observed.

\subsection{Constancy and discrepancy of activity along time}\label{constDisc}
One remarkable feature from the analysis is that the activity 
along time is practically the same for all lists. Circular mean and standard deviation
for each time scale is given in Table~\ref{tab:circ}.

\begin{table*}[t]
\scriptsize
\caption{Circular mean and standard deviation of sent messages for each time scale. The Appendix~\ref{ap:circular} exposes the theoretical background considered.}
\begin{center}
    \begin{tabular}{ |l|| c|c|}
        \hline
scale & mean & standard deviation \\ \hline
seconds & - & - \\ \hline
minutes & - & - \\ \hline
    \end{tabular}
\end{center}
\label{ano}
\end{table*}

\subsubsection{Seconds and minutes}
Messages were slightly more evenly distributed in all lists than in simulations using uniform distribution\footnote{Numpy version 1.6.1, ``random.randint'' function, was used for simulations, algorithms in \url{https://pypi.python.org/pypi/gmane}.}: for both seconds and minutes $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$ in the lists, while Simulations reach these values but have in average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$.
Therefore, the incidence of messages at each second of a minute and at each minute of an hour was considered uniform, i.e. no trend was detected and the pattern is the uniformity.

@@

\subsection{Scalable fat-tail structure: constancy of membership fractions in the Erd\"os sectors}\label{subsec:pih}
There is a concentration of hub activity and of vertex with few connections. Table~\ref{autores} is dedicated to exposing this well known and expected distribution of activity among participants.

@@

Regarding topological aspects of interaction networks, further work should inspect other measures in each of the three connective sectors: hubs, intermediary and periphery.
Observance of attributes with greater contribution to principal components of LDA should reveal best chances to present these three sections as clusters in the network measurements space. Another possibility, specially for a brute-force characterization of such sectors, is to remove vertices with degree close to $k_\Leftarrow$ or $k_\Rightarrow$ depicted in figure~\ref{fig:setores}. The subtraction $\widetilde{P}(k)-P(k)$ (see Section~\ref{subsec:typ}) results in two positive clusters for periphery and hubs, and a negative cluster for intermediary vertices. This might support classification of the three sectors by clustering, a more traditional approach to classification.

@@

\subsection{Third party libraries and software}
The programming framework used 
is mainly Python-based, with emphasis on usual
scientific tools. More specifically,
scripts where written for 2.7.3 version of Python,
with the following third party libraries: Numpy, Pylab/Matplotlib, NetworkX, IGraph.
Behind the scenes, Graphviz is accessed via PyGraphviz to make network drawings.
\section{Circular statistics}\label{ap:circular}
\section{Typologies}\label{ap:typ}
There are other ways to split and characterize networks. To point a common example, the center of the network is defined as all the nodes whose maximum distance to any other node is the radius (the radius is the minimum maximum distance to all vertices , i.e. the radius is the minimum eccentricity). 
In the same framework, the periphery (as opposed to the center) consists of the nodes whose maximum distance to any node is the diameter (diameter being the maximum geodesic on the network). Accordingly, the intermediary sector can be defined as the nodes that are not in the center or in the periphery. Interestingly, in the email networks analyzed, with such criteria, the center can often be a factor of 4 times larger than the periphery and the intermediary group often exceeds 90\% of the nodes~\cite{networkx}.

Models of human dynamics can be used to predict and classify activity. In this case, human activity is commonly considered a Poisson process, as a consequence of the randomly distributed events in time. Even so, evidence-based models suggests that human activity patterns follow non-Poisson statistics, characterized by a long tail of inactivity with bursts of rapidly occurring events~\cite{barabasiHumanDyn,barabasiPhone}.

\section{Tables}\label{sectables}
\clearpage
\subsection{PCA tables}\label{sec:pcat}
